Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-02-12 ltx/data_paper/DataPaper.tex
small edit

 p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&= \mid\boldsymbol\Sigma_{\epsilon}\mid^{-\frac{1}{2}}  \mathrm{exp}\left({-\frac{1}{2}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)^\top\boldsymbol\Sigma_{\epsilon}^{-1}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)}\right)\\
p\left(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d\right)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}}\mathrm{exp}\left(-\frac{1}{2}\left(\mathbf x_{t+1}-q\left(\mathbf  x_t\right)\boldsymbol\theta-\xi  \mathbf x_t\right)^\top\boldsymbol\Sigma_e^{-1}\left(\mathbf x_{t+1}-q\left( \mathbf x_t\right)\boldsymbol\theta-\xi \mathbf  x_t\right)\right)

@@

Substituting  \ref{eq:Qfunctiontrace1} and \ref{eq:Qfunctiontrace2} into \ref{eq:QIntermsofJointDist} and taking the expectation of the log likelihood function for all time instants gives the lower-bound that needs to be maximised to gain optimal parameter estimates,
\begin{align}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&= -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right]\boldsymbol\theta
 \nonumber \\
&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \boldsymbol\theta
\end{align}
It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right]$. By choosing
\begin{align}
	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\\
\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right] \label{eq:Xi3} \\	
\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right] \label{eq:Xi4}\\
 \Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \label{eq:Xi5}
 \end{align}
equation \ref{eq:Voldermont} can be written as
\begin{align}\label{eq:VoldermontwitXivariables}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 
 \nonumber \\
&+\frac{2}{\sigma_d^2}\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_{5} \boldsymbol\theta.
\end{align}

@@

Computing unknown variables, $\boldsymbol\Theta$, in the maximization step requires calculation of covariance and  cross-covariance terms, including both linear, $\Xi_0$, $\Xi_1$ and $\Xi_2$ and non-linear terms, $\Xi_3$, $\Xi_4$ and  $\Xi_5$. Finding linear $\Xi$-variables is straightforward as they can be obtained using smoother outputs, $\hat{\mathbf x}_t $, $\mathbf P_t$ and $\mathbf M_t$ using the following expressions \cite{Shumway2000} (see supporting information for details)
\begin{align}
	\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0}, \\
\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1}, \\
\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
 \end{align}
However calculation of non-linear $\Xi$-variables required for the M-step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the calculations of the $\Xi_3$, $\Xi_4$ and $\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the estimated state, $\hat{\mathbf x}_t$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the estimated state $\hat{\mathbf{x}}_t$ 
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'
\end{equation}
Define
\begin{equation}
	e(x) = q(x) - f(x) \nonumber
\end{equation}

Then we can write f(x) = q(x) + e(x) and use this to show that the error in the estimate is proportional to e(x).

>>>>

 p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&= \mid\boldsymbol\Sigma_{\epsilon}\mid^{-\frac{1}{2}}  \exp\left(-\frac{1}{2}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\boldsymbol\Sigma_{\epsilon}^{-1}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})\right) \\
p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}} \exp \left(-\frac{1}{2}(\mathbf x_{t+1}-q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t) \right)

@@

Substituting  \ref{eq:Qfunctiontrace1} and \ref{eq:Qfunctiontrace2} into \ref{eq:QIntermsofJointDist} and taking the expectation of the log likelihood function for all time instants gives the lower-bound that needs to be maximised to gain optimal parameter estimates,
\begin{align}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&= -(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right] \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace +\frac{2}{\sigma_d^2}\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right]\boldsymbol\theta
 \nonumber \\
&-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right]\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \boldsymbol\theta
\end{align}
It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right]$, ref(Shumway). By choosing
\begin{align}
	\Xi_{0}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]\label{eq:defofXi0} \\
\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\\
\Xi_{3}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}q( \mathbf x_t)\right] \label{eq:Xi3} \\	
\Xi_{4}&= \mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}  q^\top(\mathbf  x_t)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)\right] \label{eq:Xi4}\\
 \Xi_{5}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf x_t)  \right] \label{eq:Xi5}
 \end{align}
equation \ref{eq:Voldermont} can be written as
\begin{align}\label{eq:VoldermontwitXivariables}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-(T-1)n_y\ln \sigma_{\epsilon}^2-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&-(T-1)n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace \Xi_{0} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\frac{2\xi}{\sigma_d^2} \mathrm{tr}\left\lbrace \Xi_{1} \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -  \frac{\xi^2}{\sigma_d^2}\mathrm{tr} \left\lbrace\Xi_{2}\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace 
 \nonumber \\
&+\frac{2}{\sigma_d^2}\Xi_{3}\boldsymbol\theta-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top \Xi_{4}\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \Xi_{5} \boldsymbol\theta.
\end{align}

@@

Computing unknown variables, $\boldsymbol\Theta$, in the maximization step requires calculation of covariance and  cross-covariance terms, including both linear, $\Xi_0$, $\Xi_1$ and $\Xi_2$ and non-linear terms, $\Xi_3$, $\Xi_4$ and  $\Xi_5$. Finding linear $\Xi$-variables is straightforward as they can be obtained using smoother outputs, $\hat{\mathbf x}_t $, $\mathbf P_t$ and $\mathbf M_t$ using the following expressions \cite{Shumway2000} (see supporting information for details)
\begin{align}
	\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}+\mathbf{\hat x}_{t+1}\mathbf{\hat x}_{t+1}^\top\right)\label{eq:defofXi0}, \\
\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:defofXi1}, \\
\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
 \end{align}
However calculation of non-linear $\Xi$-variables required for the M-step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$ at each time instant which is to be used in the calculations of the $\Xi_3$, $\Xi_4$ and $\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the estimated state, $\hat{\mathbf x}_t$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ around the estimated state $\hat{\mathbf{x}}_t$ 
\begin{equation}\label{eq:qTaylor}
 q_L(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t\right) d\mathbf{r}'
\end{equation}
Define
\begin{equation}
	e(x) = q(x) - q_L(x) \nonumber
\end{equation}

Then we can write $q(x) = q_L(x) + e(x)$ and use this to show that the error in the estimate is proportional to $e(x)$.
